DynaProg for Scala

### 2013-01-02 docs/report/problems.tex
Renamed roadmap into report.
Nits.

% ------------------------------------------------------------------------------------------------
% Recurrence visualization helpers
\newcommand\Cd[3][0,-1]{\put(#2){\put(.5,.5){\circle*{.3}}\put(.5,.5){\linethickness{1.5pt}\vector(#1){#3}}}} % dependency [dx,dy]{x,y}{len}
\def\Cg#1{\put(#1){\color{lightgray}\put(0,0){\polygon*(0,0)(0,1)(1,1)(1,0)}}} % grayed cell (not to store
\def\Cz#1{\put(#1){\put(0,.35){\parbox{1\unitlength}{\centering\bf 0}}}} % zero-init cell
\def\Cm{\put(6.5,4.5){\circle*{.4}}\multiput(0,0)(1,0){9}{\line(0,1){8}}\multiput(0,0)(0,1){9}{\line(1,0){8}}} % matrix base
\def\Cfl#1{#1{0,6}#1{0,5}#1{1,5}#1{0,4}#1{1,4}#1{2,4}#1{0,3}#1{1,3}#1{2,3}#1{3,3}#1{0,2}#1{1,2}#1{2,2}#1{3,2}#1{4,2}
	#1{0,1}#1{1,1}#1{2,1}#1{3,1}#1{4,1}#1{5,1}#1{0,0}#1{1,0}#1{2,0}#1{3,0}#1{4,0}#1{5,0}#1{6,0}} % triangular lower (function)
\def\Cfd#1{#1{0,7}#1{1,6}#1{2,5}#1{3,4}#1{4,3}#1{5,2}#1{6,1}#1{7,0}} % main diagonal
\def\Cfdu#1{#1{7,1}#1{6,2}#1{5,3}#1{4,4}#1{3,5}#1{2,6}#1{1,7}} % upper diagonal

@@

@@

\item Backtracking: starts from the cell $M_{(m,n)}$ and stops at the first cell containing a $0$.
\item Visualization: by convention, we put the longest string vertically ($m\ge n$):
\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
	\put(-.5,7.5){S}\put(-.35,7.4){\linethickness{1pt}\vector(0,-1){2}}
	\put(.2,8.2){T}\put(.8,8.4){\linethickness{1pt}\vector(1,0){2}}
	\Cz{0,0}\Cz{0,1}\Cz{0,2}\Cz{0,3}\Cz{0,4}\Cz{0,5}\Cz{0,6}\Cz{0,7}
	\Cz{1,7}\Cz{2,7}\Cz{3,7}\Cz{4,7}\Cz{5,7}\Cz{6,7}\Cz{7,7}
	\Cd{6,5}{0.8}
	\Cd[1,0]{5,4}{0.8}
	\Cd[1,-1]{5,5}{0.8}
\Cm\end{picture}\end{center}

@@

Although the latter notation seems more explicit, it introduces non-serial dependencies that the former set of recurrences is free of. So we need to implement the former rules whose kernel is 
\[ [M;E;F]_{(i,j)} = f_{\rm kernel} ( [M;E]_{(i,j-1)}, [M;F]_{(i-1,j)}, M_{(i-1,j-1)} ) \]
Notice that this recurrence is very similar to \nameref{sswat} except that we propagate 3 values ($M,E,F$) instead of a single one ($M$). Also notice that it is possible to propagate $E$ and $F$ inside a resp. horizontal and vertical wavefront.

\item Backtracking, visualization and optimizations are the same as in \nameref{sswat}
\ole

@@

\item Backtracking: similar to \nameref{sswat} except that you can jump of $k$ cells along the rows or along the columns.
\item Visualization:
	\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
		\put(-.5,7.5){S}\put(-.35,7.4){\linethickness{1pt}\vector(0,-1){2}}
		\put(.2,8.2){T}\put(.8,8.4){\linethickness{1pt}\vector(1,0){2}}
	\Cz{0,0}\Cz{0,1}\Cz{0,2}\Cz{0,3}\Cz{0,4}\Cz{0,5}\Cz{0,6}\Cz{0,7}
	\Cz{1,7}\Cz{2,7}\Cz{3,7}\Cz{4,7}\Cz{5,7}\Cz{6,7}\Cz{7,7}
		\Cd[0,-1]{6,7}{2.8}\Cd[0,-1]{6,6}{1.8}\Cd{6,5}{0.8}
		\Cd[1,0]{0,4}{5.8}\Cd[1,0]{1,4}{4.8}\Cd[1,0]{2,4}{3.8}\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
		\Cd[1,-1]{5,5}{0.8}
	\Cm\end{picture}\end{center}

@@

% ----------------------------------------------
\newpage
\subsubsection{Convex polygon triangulation}\ol
\item Problem: triangulating a polygon of $n$ vertices with least total cost for added edges. We denote the cost of adding an edge between the  pair of edges $i,j$ by $S_{(i,j)}$, Where $S_{n \times n}$ is a symmetric matrix (can be stored as a triangular matrix with 0 diagonal that can be omitted), hence $|S|=\tfrac{n(n-1)}{2}=N$.
\item Matrices: $M_{(n+1)\times (n+1)}, B_{(n+1)\times (n+1)}$ \textit{indices denote <<first vertex, last vertex>>}.\\ Upper triangular matrices including main diagonal. Note that the vertex $n$ is the same as the vertex $0$ due to the cyclic nature of the problem.
\item Alphabets:\ul
	\item Input: $\Sigma(S_{(i,j)})=\{0..m\}$ with $m=\max_{i,j} S_{(i,j)}$ determined at runtime\footnote{We need to have statistics about $S$, this is where dynamic compilation might play a role}.
	\item Cost matrix: $\Sigma(M)=\{0..z\}$ with $z = m \cdot (n-2)$ (a triangulation of a polygon of $n$ edges adds at most $n-2$ edges).
	\item Backtrack matrix: $\Sigma(B)=\{stop, 0..n\}$ (index of intermediate edge)
	\ule
\item Initialization: $M_{(i,i)}=M_{(i,i+1)}=0, B_{(i,i)}=B_{(i,i+1)}=stop \quad\forall i$
\item Recurrence: \[M_{(i,j)}=\left\{ S(i,j) + \max_{i<k<j}M_{(i,k)}+M_{(k,j)} \,\,\rule[-.75em]{.5pt}{2em}\,\,  k \right\} = B_{(i,j)} \]
	Intuition: triangulate the partial polygon $(i,..j)$ recursively. For the last triangle of the polygon consider 3 cases:\ul
	\item Given 2 triangulations $(1..k)$ and $(k..n)$, we close the polygon with $\bigtriangleup(1,k,n)$
	\item Given a triangulation $(1..n-1)$, we close the polygon with $\bigtriangleup(1,n-1,n)$
	\item Given a triangulation $(2..n)$, we close the polygon with $\bigtriangleup(1,2,n)$
	\ule
	Since the edge to close the last triangle is already part of the polygon, its cost is 0.
\item Backtracking: Add the edges in the set given by the set ${\rm BT}(B_{(0,n)})$ where
	\[{\rm BT}(B_{(i,j)}=k) \mapsto \left\{\begin{array}{ll} \{\} & \text{if } k=stop \\
		 \{(i,j)\} \cup {\rm BT}(B_{(i,k)}) \cup {\rm BT}(B_{(k,j)}) & \text{otherwise} \end{array}\right.\]
\item Visualization:
\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
	\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
	\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
	\Cfl{\Cg}
	\Cfd{\Cz}\Cfdu{\Cz}
	\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
	\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
	\put(3.5,4.5){\line(3,-1){3}}
	\put(4.5,4.5){\line(2,-2){2}}
	\put(5.5,4.5){\line(1,-3){1}}
	%\Cd[1,-1]{5,2}{0.8}
	\Cm %\Cmlong
\end{picture}\end{center}

\item Optimizations: we need the cost of edges between contiguous polygon vertices to have a cost of 0, so that we do not need to handle special cases in the DP program (i.e. existing edges cannot be added).
\ole

% ----------------------------------------------
\newpage
\subsubsection{Matrix chain multiplication}\ol
\item Problem: find an optimal parenthesizing of the multiplication of $n$ matrices $A_i$. Each matrix $A_i$ is of dimension $r_i \times c_i$ and $c_i=r_{i+1} \forall i$. \textit{<<r=rows, c=columns>>}
\item Matrices: $M_{n \times n}, B_{n \times n}$ \textit{(first, last matrix)}
\item Alphabets:\ul
	\item Input: matrix $A_i$ size is defined as pairs of integers $(r_i,c_i)$.
	\item Cost matrix: $\Sigma(M)= 1..z$ with $z\le n\cdot \big[ \max_i(r_i,c_i) \big]^3 $.
	\item Backtrack matrix: $\Sigma(B)=\{stop\} \cup \{0..n\}$.
	\ule
\item Initialization:\ul
	\item Cost matrix: $M_{(i,i)}=0$.
	\item Backtrack matrix: $B_{(i,i)}=stop$.
	\ule
\item Recurrence: $c_k=r_{k+1}$
	\[M_{(i,j)}=\min_{i\le k<j}\left\{\begin{array}{l|l}
		M_{(i,k)}+M_{(k+1,j)}+r_i \cdot c_k \cdot c_j & k
	\end{array}\right\}=B_{(i,j)} \]
\item Backtracking: Start at $B_{(0,n-1)}$. Use the following recursive function for parenthesizing
	\[{\rm BT}(B_{(i,j)}=k) \mapsto \left\{\begin{array}{ll} A_i & \text{if } k=stop \\
		\Big( {\rm BT}(B_{(i,k)}) \Big) \cdot \Big( {\rm BT}(B_{(k+1,j)}) \Big) & \text{otherwise} \end{array}\right.\]

\item Visualization:
	\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
		\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
		\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
		\Cfl{\Cg}\Cfd{\Cz}
		\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
		\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
		\put(3.5,4.5){\line(3,-1){3}}\put(4.5,4.5){\line(2,-2){2}}\put(5.5,4.5){\line(1,-3){1}}
	\Cm\end{picture}\end{center}

@@

% ----------------------------------------------
\newpage
\subsubsection{Nussinov algorithm}\ol
\item Problem: folding a RNA string $S$ over itself $|S|=n$.
\item Matrices: $M_{n\times n}, B_{n \times n}$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\{a,c,g,u\}$.
	\item Cost matrix: $\Sigma(M)=\{0..n\}$
	\item Backtrack matrix: $\Sigma(B)=\{stop,D,1..n\}$
	\ule
\item Initialization: \ul
	\item Cost matrix: $ M_{(i,i)}=M_{(i,i+1)}=0$
	\item Backtrack matrix: $B_{(i,i)}=B_{(i,i+1)}=stop$
	\ule
\item Recurrences:
	\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
		M_{(i+1,j-1)}+\omega(i,j) & D\\
		\max_{i\le k<j}M_{(i,k)}+M_{(k+1,j)} & k
	\end{array}\right\} = B_{(i,j)} \]
	With $\omega(i,j)=1$ if $i,j$ are complementary. 0 otherwise.
\item Backtracking: Start the backtracking in $B_{(0,n-1)}$ and go backward. The backtracking is very similar to that of the matrix multiplication, except that we also introduce the diagonal matching.
\item Visualization:
	\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
		\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
		\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
		\Cfl{\Cg}\Cfd{\Cz}\Cfdu{\Cz}
		\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
		\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
		\Cd[1,1]{5,3}{0.8}
		\put(3.5,4.5){\line(3,-1){3}}\put(4.5,4.5){\line(2,-2){2}}\put(5.5,4.5){\line(1,-3){1}}
	\Cm\end{picture}\end{center}

@@

\item Visualizations\footnote{Reproductions of the illustrations from \cite{para_dprec} pp.148,149}:
\begin{center}\includegraphics[width=8cm]{inc/zuker_rec.pdf}\end{center}
The recurrence consists of two non-serial dependencies as in \textit{\nameref{aswat}} plus a bounded 2-dimensional dependency for bulges.\\[12pt]
Since this problem is non-trivial to understand from the recurrences, we propose an additional illustration of a RNA chain folded according to the Zuker folding algorithm.
\begin{center}
\includegraphics[width=10cm]{inc/zuker_struct.pdf} \\
\small An example of an RNA folded into a secondary structure. Types of structural features modeled by the Zuker folding algorithm include: dangling ends (1), internal loop (11), stack (23), multi-loop (47), bulge (68) and hairpin loop (78).
\end{center}

@@

% ------------------------------------------------------------------------------------------------
\newpage
\subsection{Related problems}
The goal of this section is to demonstrate that the problems described above are very similar or encompass a significant part of the common dynamic programming problems. (There is an hyperlink on the problems name to their detailed description).

\subsubsection{Serial problems}
\begin{tabular}{llcc} \toprule
\bf Problem & \bf Shape & \bf Matrices & \bf Wavefront \\ \midrule
Smith-Waterman \footnotesize simple & rectangle & 1 & -- \\
Smith-Waterman \footnotesize affine gap extension & rectangle & 3 & (can replace 2 matrices) \\
\href{http://en.wikipedia.org/wiki/Needleman-Wunsch_algorithm}{Needleman-Wunsch} & rectangle & 1 & -- \\

\href{http://en.wikipedia.org/wiki/Dynamic_programming#Checkerboard}{Checkerboard} & rectangle & 1 & -- \\
\href{http://en.wikipedia.org/wiki/Longest_common_subsequence_problem\#Code_for_the_dynamic_programming_solution}{Longest common subsequence} & rectangle & 1 & -- \\
\href{http://en.wikipedia.org/wiki/Longest_common_substring_problem\#Pseudocode}{Longest common substring} & triangle & 1 & -- \\
\href{http://en.wikipedia.org/wiki/Levenshtein_distance\#Computing_Levenshtein_distance}{Levenshtein distance} & rectangle & 1 & -- \\
\href{http://en.wikipedia.org/wiki/De_Boor's_algorithm}{De Boor} \footnotesize evaluating B-spline curves & rectangle & 1 & -- \\
\end{tabular}

\subsubsection{Non-serial problems}
\begin{tabular}{llcc} \toprule
\bf Problem & \bf Shape & \bf Matrices & \bf Wavefront \\ \midrule
Smith-Waterman \footnotesize arbitrary gap cost & rectangle & 1 & -- \\
Convex polygon triangulation & triangle & 1 & -- \\
Matrix chain multiplication & triangle & 1 & -- \\
Nussinov & triangle & 1 & -- \\
Zuker folding & triangle & 3 & -- \\
\href{http://en.wikipedia.org/wiki/CYK_algorithm}{CYK} \footnotesize Cocke-Younger-Kasami & triangle & \#rules & -- \\
\href{http://en.wikipedia.org/wiki/Knapsack_problem#Dynamic_programming}{Knapsack} \footnotesize (pseudo-polynomial) & rectangle & 1 & --\\
\end{tabular}

@@

\subsubsection{Conclusion}
In the rest of the project, we use a different description of the problems that is based on ADP \cite{adp}, which is more convenient but does not share much with the above description (even though ultimately the executed computations are very similar). Although not of immediate use, the description of the above problem and ad-hoc CUDA implementation of three of them (Smith-Waterman with arbitrary gap cost, Matrix chain multiplication and Convex polygon triangulation) helped us to understand:\ol
\item There is a difference between dynamic programming as seen in algorithmic classes and their concrete implementation, mainly because special care must be taken for correct indices and preventing off-by-one errors.
\item Problems can be classified in two categories: single track (input) and two-tracks (2 input sequences). Most of the common dynamic programming problems fall in these two categories.
\item Sometimes matrices are initially padded with zeroes (or initial value), although this might be ignored at algorithm design, care must be taken for these special values and their inclusion in the matrix should be decided according to the complexity of the recurrence formula.
\item Incidentally, we proposed a cyclic variant of the convex polygon triangulation, which uses a parallelogram matrix instead of a triangular one (upper right triangle extended by a lower triangle on the right). Unfortunately, this proved to be based on an erroneous recurrence relation analysis, which could be reduced to the more natural triangular case. It is not impossible that some problems do require such special structure, but we did not find any problem that could justify such structure. 
\ole

>>>>

% ------------------------------------------------------------------------------------------------
% Recurrence visualization helpers

\newcommand\Cd[3][0,-1]{\put(#2){\put(.5,.5){\circle*{.3}}\put(.5,.5){\linethickness{1.5pt}\vector(#1){#3}}}} % dependency [dx,dy]{x,y}{len}
\def\Cg#1{\put(#1){\color{lightgray}\put(0,0){\polygon*(0,0)(0,1)(1,1)(1,0)}}} % grayed cell (not to store
\def\Cz#1{\put(#1){\put(0,.35){\parbox{1\unitlength}{\centering\bf 0}}}} % zero-init cell
\def\Cm{\put(6.5,4.5){\circle*{.4}}\multiput(0,0)(1,0){9}{\line(0,1){8}}\multiput(0,0)(0,1){9}{\line(1,0){8}}} % matrix base
\def\Cfl#1{#1{0,6}#1{0,5}#1{1,5}#1{0,4}#1{1,4}#1{2,4}#1{0,3}#1{1,3}#1{2,3}#1{3,3}#1{0,2}#1{1,2}#1{2,2}#1{3,2}#1{4,2}
	#1{0,1}#1{1,1}#1{2,1}#1{3,1}#1{4,1}#1{5,1}#1{0,0}#1{1,0}#1{2,0}#1{3,0}#1{4,0}#1{5,0}#1{6,0}} % triangular lower (function)
\def\Cfd#1{#1{0,7}#1{1,6}#1{2,5}#1{3,4}#1{4,3}#1{5,2}#1{6,1}#1{7,0}} % main diagonal
\def\Cfdu#1{#1{7,1}#1{6,2}#1{5,3}#1{4,4}#1{3,5}#1{2,6}#1{1,7}} % upper diagonal

@@

\def\CFig#1#2{\begin{figure}[H]\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9) #2 \Cm\end{picture}\end{center}\caption{#1}\end{figure}}

@@

\item Backtracking: starts from the cell $M_{(m,n)}$ and stops at the first cell containing a $0$.
\item Visualization: by convention, we put the longest string vertically ($m\ge n$):
\CFig{Smith-Waterman (affine gap cost) dependencies (serial)}{
	\put(-.5,7.5){S}\put(-.35,7.4){\linethickness{1pt}\vector(0,-1){2}}
	\put(.2,8.2){T}\put(.8,8.4){\linethickness{1pt}\vector(1,0){2}}
	\Cz{0,0}\Cz{0,1}\Cz{0,2}\Cz{0,3}\Cz{0,4}\Cz{0,5}\Cz{0,6}\Cz{0,7}
	\Cz{1,7}\Cz{2,7}\Cz{3,7}\Cz{4,7}\Cz{5,7}\Cz{6,7}\Cz{7,7}
	\Cd{6,5}{0.8}\Cd[1,0]{5,4}{0.8}\Cd[1,-1]{5,5}{0.8}
}

@@

Although the latter notation seems more explicit, it introduces non-serial dependencies that the former set of recurrences is free of. So we need to implement the former rules whose kernel is 
\[ [M;E;F]_{(i,j)} = f_{\rm kernel} ( [M;E]_{(i,j-1)}, [M;F]_{(i-1,j)}, M_{(i-1,j-1)} ) \]

\item Backtracking: similar to \nameref{sswat}
\item Visualization: same as \nameref{sswat}
\item Optimizations: Notice that this recurrence is very similar to \nameref{sswat} except that we propagate 3 values ($M,E,F$) instead of a single one ($M$). Also notice that it is possible to propagate $E$ and $F$ inside a respectively horizontal and vertical wavefront, hence removing the need of two additional matrices.
\ole

@@

\item Backtracking: similar to \nameref{sswat} except that you can jump of $k$ cells along the rows or along the columns.
\item Visualization:
\CFig{Smith-Waterman (arbitrary gap cost) dependencies}{
	\put(-.5,7.5){S}\put(-.35,7.4){\linethickness{1pt}\vector(0,-1){2}}
	\put(.2,8.2){T}\put(.8,8.4){\linethickness{1pt}\vector(1,0){2}}
	\Cz{0,0}\Cz{0,1}\Cz{0,2}\Cz{0,3}\Cz{0,4}\Cz{0,5}\Cz{0,6}\Cz{0,7}
	\Cz{1,7}\Cz{2,7}\Cz{3,7}\Cz{4,7}\Cz{5,7}\Cz{6,7}\Cz{7,7}
	\Cd[0,-1]{6,7}{2.8}\Cd[0,-1]{6,6}{1.8}\Cd{6,5}{0.8}
	\Cd[1,0]{0,4}{5.8}\Cd[1,0]{1,4}{4.8}\Cd[1,0]{2,4}{3.8}\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
	\Cd[1,-1]{5,5}{0.8}
}

@@

% ----------------------------------------------
\newpage
\subsubsection{Convex polygon triangulation}\ol
\item Problem: triangulating a polygon of $n$ vertices with least total cost for added edges. We denote the cost of adding an edge between the  pair of edges $i,j$ by $S_{(i,j)}$, Where $S_{n \times n}$ is a symmetric matrix (can be stored as a triangular matrix with 0 diagonal that can be omitted), hence $|S|=\tfrac{n(n-1)}{2}=N$.
\item Matrices: $M_{(n+1)\times (n+1)}, B_{(n+1)\times (n+1)}$ \textit{indices denote <<first vertex, last vertex>>}.\\ Upper triangular matrices including main diagonal. Note that the vertex $n$ is the same as the vertex $0$ due to the cyclic nature of the problem.
\item Alphabets:\ul
	\item Input: $\Sigma(S_{(i,j)})=\{0..m\}$ with $m=\max_{i,j} S_{(i,j)}$ determined at runtime\footnote{We need to have statistics about $S$, this is where dynamic compilation might play a role}.
	\item Cost matrix: $\Sigma(M)=\{0..z\}$ with $z = m \cdot (n-2)$ (a triangulation of a polygon of $n$ edges adds at most $n-2$ edges).
	\item Backtrack matrix: $\Sigma(B)=\{stop, 0..n\}$ (index of intermediate edge)
	\ule
\item Initialization: $M_{(i,i)}=M_{(i,i+1)}=0, B_{(i,i)}=B_{(i,i+1)}=stop \quad\forall i$
\item Recurrence: \[M_{(i,j)}=\left\{ S(i,j) + \max_{i<k<j}M_{(i,k)}+M_{(k,j)} \,\,\rule[-.75em]{.5pt}{2em}\,\,  k \right\} = B_{(i,j)} \]
	Intuition: triangulate the partial polygon $(i,..j)$ recursively. 3 cases for the last triangle:\ul
	\item Given 2 triangulations $(1..k)$ and $(k..n)$, we close the polygon with $\bigtriangleup(1,k,n)$
	\item Given a triangulation $(1..n-1)$, we close the polygon with $\bigtriangleup(1,n-1,n)$
	\item Given a triangulation $(2..n)$, we close the polygon with $\bigtriangleup(1,2,n)$
	\ule
	Since the edge to close the last triangle is already part of the polygon, its cost is 0.
\item Backtracking: Add the edges in the set given by the set ${\rm BT}(B_{(0,n)})$ where
	\[{\rm BT}(B_{(i,j)}=k) \mapsto \left\{\begin{array}{ll} \{\} & \text{if } k=stop \\
		 \{(i,j)\} \cup {\rm BT}(B_{(i,k)}) \cup {\rm BT}(B_{(k,j)}) & \text{otherwise} \end{array}\right.\]
\item Visualization:
\CFig{Convex polygon triangulation dependencies}{
	\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
	\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
	\Cfl{\Cg}\Cfd{\Cz}\Cfdu{\Cz}
	\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
	\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
	\put(3.5,4.5){\line(3,-1){3}}
	\put(4.5,4.5){\line(2,-2){2}}
	\put(5.5,4.5){\line(1,-3){1}}
}

\item Optimizations: if the cost of edges between contiguous vertices is 0, we do not need to handle special cases in the DP program (i.e. existing edges cannot be added).
\ole

% ----------------------------------------------
\newpage
\subsubsection{Matrix chain multiplication}\label{mat_mult_plain}\ol
\item Problem: find an optimal parenthesizing of the multiplication of $n$ matrices $A_i$. Each matrix $A_i$ is of dimension $r_i \times c_i$ and $c_i=r_{i+1} \forall i$. \textit{<<r=rows, c=columns>>}
\item Matrices: $M_{n \times n}, B_{n \times n}$ \textit{(first, last matrix)}
\item Alphabets:\ul
	\item Input: matrix $A_i$ size is defined as pairs of integers $(r_i,c_i)$.
	\item Cost matrix: $\Sigma(M)= 1..z$ with $z\le n\cdot \big[ \max_i(r_i,c_i) \big]^3 $.
	\item Backtrack matrix: $\Sigma(B)=\{stop\} \cup \{0..n\}$.
	\ule
\item Initialization:\ul
	\item Cost matrix: $M_{(i,i)}=0$.
	\item Backtrack matrix: $B_{(i,i)}=stop$.
	\ule
\item Recurrence: $c_k=r_{k+1}$
	\[M_{(i,j)}=\min_{i\le k<j}\left\{\begin{array}{l|l}
		M_{(i,k)}+M_{(k+1,j)}+r_i \cdot c_k \cdot c_j & k
	\end{array}\right\}=B_{(i,j)} \]
\item Backtracking: Start at $B_{(0,n-1)}$. Use the following recursive function for parenthesizing
	\[{\rm BT}(B_{(i,j)}=k) \mapsto \left\{\begin{array}{ll} A_i & \text{if } k=stop \\
		\Big( {\rm BT}(B_{(i,k)}) \Big) \cdot \Big( {\rm BT}(B_{(k+1,j)}) \Big) & \text{otherwise} \end{array}\right.\]

\item Visualization:
\CFig{Matrix chain multiplication dependencies}{
	\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
	\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
	\Cfl{\Cg}\Cfd{\Cz}
	\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
	\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
	\put(3.5,4.5){\line(3,-1){3}}\put(4.5,4.5){\line(2,-2){2}}\put(5.5,4.5){\line(1,-3){1}}
}

@@

% ----------------------------------------------
\newpage
\subsubsection{Nussinov algorithm}\ol
\item Problem: folding a RNA string $S$ over itself $|S|=n$.
\item Matrices: $M_{n\times n}, B_{n \times n}$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\{a,c,g,u\}$.
	\item Cost matrix: $\Sigma(M)=\{0..n\}$
	\item Backtrack matrix: $\Sigma(B)=\{stop,D,1..n\}$
	\ule
\item Initialization: \ul
	\item Cost matrix: $ M_{(i,i)}=M_{(i,i+1)}=0$
	\item Backtrack matrix: $B_{(i,i)}=B_{(i,i+1)}=stop$
	\ule
\item Recurrences:
	\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
		M_{(i+1,j-1)}+\omega(i,j) & D\\
		\max_{i\le k<j}M_{(i,k)}+M_{(k+1,j)} & k
	\end{array}\right\} = B_{(i,j)} \]
	With $\omega(i,j)=1$ if $i,j$ are complementary. 0 otherwise.
\item Backtracking: Start the backtracking in $B_{(0,n-1)}$ and go backward. The backtracking is very similar to that of the matrix multiplication, except that we also introduce the diagonal matching.
\item Visualization:
\CFig{Nussinov dependencies}{
	\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
	\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
	\Cfl{\Cg}\Cfd{\Cz}\Cfdu{\Cz}
	\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
	\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
	\Cd[1,1]{5,3}{0.8}
	\put(3.5,4.5){\line(3,-1){3}}\put(4.5,4.5){\line(2,-2){2}}\put(5.5,4.5){\line(1,-3){1}}
}

@@

\item Visualizations\footnote{Reproductions of the illustrations from \cite{para_dprec} pp.148,149}:
\begin{figure}[H]\begin{center}\includegraphics[width=8cm]{inc/zuker_rec.pdf}\end{center}\caption{Zuker folding dependencies}\end{figure}

The recurrence consists of two non-serial dependencies as in \textit{\nameref{aswat}} plus a bounded 2-dimensional dependency for bulges.\\[12pt]
Since this problem is non-trivial to understand from the recurrences, we propose an additional illustration of a RNA chain folded according to the Zuker folding algorithm.
\begin{figure}[H]\begin{center}
\includegraphics[width=10cm]{inc/zuker_struct.pdf} \\
\small Types of structural features modeled by the Zuker folding algorithm include: dangling ends (1), \\ internal loop (11), stack (23), multi-loop (47), bulge (68) and hairpin loop (78).\end{center}\caption{An example of an RNA folded into a secondary structure}\end{figure}

@@

% ------------------------------------------------------------------------------------------------
\newpage
\subsection{Related problems}
The aim of this section is to demonstrate that the problems described above are very similar or encompass a significant part of the common dynamic programming problems. (There is an hyperlink on the problems name to their detailed description).

\begin{table}[H]\begin{center}\begin{tabular}{llcc} \toprule
\bf Serial problems & \bf Shape & \bf Matrices & \bf Wavefront \\ \midrule
Smith-Waterman \footnotesize simple & rectangle & 1 & -- \\
Smith-Waterman \footnotesize affine gap extension & rectangle & 3 & (can replace 2 matrices) \\
\href{http://en.wikipedia.org/wiki/Needleman-Wunsch_algorithm}{Needleman-Wunsch} & rectangle & 1 & -- \\
\href{http://en.wikipedia.org/wiki/Dynamic_programming#Checkerboard}{Checkerboard} & rectangle & 1 & -- \\
\href{http://en.wikipedia.org/wiki/Longest_common_subsequence_problem\#Code_for_the_dynamic_programming_solution}{Longest common subsequence} & rectangle & 1 & -- \\
\href{http://en.wikipedia.org/wiki/Longest_common_substring_problem\#Pseudocode}{Longest common substring} & triangle & 1 & -- \\
\href{http://en.wikipedia.org/wiki/Levenshtein_distance\#Computing_Levenshtein_distance}{Levenshtein distance} & rectangle & 1 & -- \\
\href{http://en.wikipedia.org/wiki/De_Boor's_algorithm}{De Boor} \footnotesize evaluating B-spline curves & rectangle & 1 & -- \vspace{.4cm} \\
\toprule
\bf Non-serial problems & \bf Shape & \bf Matrices & \bf Wavefront \\ \midrule
Smith-Waterman \footnotesize arbitrary gap cost & rectangle & 1 & -- \\
Convex polygon triangulation & triangle & 1 & -- \\
Matrix chain multiplication & triangle & 1 & -- \\
Nussinov & triangle & 1 & -- \\
Zuker folding & triangle & 3 & -- \\
\href{http://en.wikipedia.org/wiki/CYK_algorithm}{CYK} \footnotesize Cocke-Younger-Kasami & triangle & \#rules & -- \\
\href{http://en.wikipedia.org/wiki/Knapsack_problem#Dynamic_programming}{Knapsack} \footnotesize (pseudo-polynomial) & rectangle & 1 & --\\
\end{tabular}\end{center}
\caption{Classification of related problems}
\end{table}

@@

\subsubsection{Conclusion}
In the rest of the project, we use a different description of the problems that is based on ADP \cite{adp}, which is more convenient but does not share much with the above description (even though ultimately the executed computations are very similar). Although not of immediate use, the description of the above problem and ad-hoc CUDA implementation of three of them (Smith-Waterman with arbitrary gap cost, Matrix chain multiplication and Convex polygon triangulation) helped us to understand:\ol
\item There is a difference between dynamic programming as seen in algorithmic classes and their concrete implementation, mainly because special care must be taken for correct indices and preventing off-by-one errors.
\item Problems can be classified in two categories: single track (input) and two-tracks (2 input sequences). Most of the common dynamic programming problems fall in these two categories.
\item Sometimes matrices are initially padded with zeroes (or initial value), although this might be ignored at algorithm design, care must be taken for these special values and their inclusion in the matrix should be decided according to the complexity of the recurrence formula.
\item Incidentally, we proposed a cyclic variant of the convex polygon triangulation, which uses a parallelogram matrix instead of a triangular one (upper right triangle extended by a lower triangle on the right). Unfortunately, this proved to be based on an erroneous recurrence relation analysis, which could be reduced to the more natural triangular case.

Although the polygon triangulation problem can be solved with a regular triangular matrix and we have not found real problem requiring a parallelogram matrix, we still present this version in the analysis part of the project, which could be justified for cyclic problems that require dynamic programming for each unfolding position (that is for all position in the circular structure, break the cycle at this position, and solve the dynamic programming problem on the resulting flattened sequence). For example, one could be interested in finding the longest subsequence verifying some property in a cycle, such that the subsequence score changes if it is rotated.

\ole
